\textcolor{sage}{Does Hospital Leadership Matter? \newline Evidence from Pay-for-Performance}

### 2024-11-06 Slides/JobTalk.tex
Updates from Overleaf

\begin{block}{\normalsize Gaps in Literature}
    \small 
    \begin{enumerate}
        \item Limited to large, publicly traded firms: reason to believe this does not extend to nonprofit setting
        \item Direct responses to policy changes are not well-understood
    \end{enumerate}
\end{block}
\end{frame}

@@

\textbf{Hospital care makes up a third of health care spending} \scriptsize (\cite{cms2024nhe}) \normalsize
\begin{itemize}
    \item A majority of hospitals are nonprofit and their objectives are not as clear as classic for-profit firms
\end{itemize}

@@

\textbf{Policies often aim to improve the value of health care delivery}

    \begin{wideitemize}
        \item It is unclear how hospital leaders view value
        \item Or whether they might play a role in achieving this goal
    \end{wideitemize}
\end{frame}

\begin{frame}{Research Question}
    \centering
    \large
    Does having executives with clinical training affect response to pay-for-performance incentives of nonprofit hospitals?
\end{frame}

@@

\textbf{Physician executives could be good or bad}

@@

\begin{wideitemize}
    \item Non-clinical reduce readmissions by blank more than clinical
    \item Non-clinical executive teams behave like profit-driven hospitals
\end{wideitemize}
\end{frame}

@@

\begin{frame}{Rest of Talk}
    Theoretical Motivation

@@

@@

@@

\begin{frame}{Nonprofit Hospital Executives}\label{nonprofithospexec}
\textbf{What do executives do?}
\begin{wideitemize}
    \item Board of directors selects executives
    \item Executives handle the day-to-day operations of the hospital \hyperlink{executivejob}{\beamergotobutton{executive job}} 
\end{wideitemize}

@@

\textbf{Executive teams}
\begin{wideitemize}
    \item Lots of variation in what executive teams look like
    \item Typically CEO and CFO, sometimes CMO 
\end{wideitemize}

Usable data on executives is rare
\end{frame}

\begin{frame}{Merge to Other Hospital Data}
    I create links between EIN and ID number from the American Hospital Association Survey (AHA) based on hospital name and location

    \vspace{7mm}
    Merge to:

    \vspace{2mm}
    \begin{itemize}
        \item AHA data for hospital demographics
        \item Hospital Compare data for readmission and mortality rates
        \item Center for Medicare and Medicaid Services (CMS) Case Mix Index file for patient complexity
        \item Health Care Cost Report Information System (HCRIS) for information on pay-for-performance rewards or penalties
    \end{itemize}
\end{frame}

\begin{frame}{Merge to Physician Data}
    I use executives that self-identify as MD

    \vspace{8mm}

    To verify, I match all executive names to database of all registered physicians and match NPI by hand
\end{frame}

    \pause

    \begin{block}{\normalsize Takeaway}
        \normalsize All hospitals in the middle of the distribution had incentive to improve quality to try to avoid penalties, which were not perfectly predictable
    \end{block}
\end{frame}

\begin{frame}{Executive Teams}
    \begin{table}[ht!]
        \centering
        \begin{tabular}[t]{lccc}
        \toprule
         & Ever & Always & Never\\
         & Clinical Exec & Clinical Exec &  Clinical Exec \\
        \midrule
        Number Executives & 4.85 & 5.25 & 3.11\\
        Frac Clinical Execs & 0.20 & 0.26 & 0\\
        Frac Int. Medicine Execs & 0.06 & 0.10 & 0.0\\
        Has a CMO & 0.53 & 0.45 & 0.15\\
        Has Clinical CEO & 0.25 & 0.11 & 0.0\\
        \addlinespace[0.6em]
        Num. Hospitals & 391 & 56 & 445\\
        \bottomrule
        \end{tabular}
    \end{table}
\end{frame}

    \includegraphics[width=.6\textwidth]{Objects/change_analysis_plot.pdf}

\begin{frame}{}
    put graph of outcomes here
\end{frame}

\begin{transitionframe}
\centering \LARGE
    \textcolor{black}{Effect of Clinical Leadership on Response}
\end{transitionframe}

\begin{frame}{Estimation}\label{est_equation}

    \vspace{5mm}\large
    \begin{equation*}
    y_{ht} = \beta \text{ (clinical training x post 2012)}_{ht} + \gamma_{h} + \delta_t + \epsilon_{ht}
    \end{equation*}
    \begin{itemize}
        \item Compare always to never clinical executives
        \item Leverage the timing of the programs
    \end{itemize}

    \vspace{8mm}

    Identifies the effect of having a clinically trained executive on readmission and mortality rates after pay-for-performance

    \begin{itemize}
        \item Assume that hospitals would have had parallel trends absent clinical executives
    \end{itemize}

    \btVFill
    \normalsize
    \hyperlink{synthdid_equation}{\beamergotobutton{weighted}} \hyperlink{match_equation}{\beamergotobutton{matching}}
\end{frame}

\begin{frame}{Readmission Rates}\label{main:readrates}
\centering
    \includegraphics[width=.45\textwidth]{Objects/read_md_nomd_synth_graph.pdf}

    \hyperlink{int_margin}{\beamergotobutton{intensive margin}}
\end{frame}

\begin{frame}{Mortality Rates}

@@

    \vspace{5mm}\large
    \begin{equation*}
    y_{ht} = \beta \text{ (for-profit x post 2012)}_{ht} + \gamma_{h} + \delta_t + \epsilon_{ht}
    \end{equation*}
    \begin{itemize}
        \item Two comparison groups: clinical and non-clinical executive teams
    \end{itemize}

@@

\begin{frame}{Results}

@@

@@

\begin{frame}{Results}

\begin{frame}{Conclusion}

\begin{frame}{Future Work}

\begin{transitionframe}
\centering \Large
    \textcolor{black}{hkagele@emory.edu}
\end{transitionframe}

>>>>

\begin{block}{\normalsize Gaps in Literature}
    \small 
    \begin{enumerate}
        \item Nonprofit firms: reason to believe these findings might not inform leadership in other settings
        \item Direct responses to policy changes
    \end{enumerate}
\end{block}
\end{frame}

@@

\begin{enumerate}
    \item \textbf{Hospital care makes up a third of health care spending} \scriptsize (\cite{cms2024nhe}) \normalsize

    \begin{itemize}
        \item Nonprofits drive the hospital industry \scriptsize (\cite{ASPE_2023}) \normalsize
    \end{itemize}

@@

    \item \textbf{Policies often aim to improve the value of health care delivery}

    \begin{wideitemize}
        \item It is unclear how hospital leaders view value
        \item Or whether they might play a role in achieving better value
    \end{wideitemize}

    \end{enumerate}
\end{frame}

\begin{frame}{Research Question}
    \centering
    \large
    Does having executives with clinical training affect response to pay-for-performance incentives of nonprofit hospitals in the US?
\end{frame}

@@

\textbf{Do physicians make good firm leaders?}

@@

\begin{wideitemize}
    \item Non-clinical reduce readmission by 2\% more than clinical
    \item Non-clinical executive teams behave like profit-driven hospitals
\end{wideitemize}
\end{frame}

@@

\begin{frame}{Rest of Talk}

@@

    Theoretical Motivation

@@

\begin{frame}{Nonprofit Hospital Executives}\label{nonprofithospexec}
\textbf{What do executives do?}

\begin{wideitemize}
    \item Board of directors selects executives
    \item Executives handle the day-to-day operations of the hospital \hyperlink{executivejob}{\beamergotobutton{executive job}} 
\end{wideitemize}

\textbf{Executive teams}

\begin{wideitemize}
    \item Lots of variation in what executive teams look like
    \item Typically CEO and CFO, sometimes CMO 
    \item I focus on the team as a whole instead of specific executive roles
\end{wideitemize}

\begin{frame}{Merge to Other Data}
    \textbf{Hospital:} link EIN and ID number from the American Hospital Association (AHA) Survey based on name and location

    \vspace{1mm}
    \begin{itemize}
        \item \underline{AHA} data for hospital demographics
        \item \underline{Hospital Compare} data for readmission and mortality rates
        \item Center for Medicare and Medicaid Services (CMS) \underline{Case Mix Index file} for patient complexity
        \item \underline{Health Care Cost Report Information System (HCRIS)} for information on pay-for-performance rewards or penalties
    \end{itemize}

    \textbf{Physician:} I use executives that self-identify as MD in the tax form data

    \begin{itemize}
        \item Manually match physician executives to \underline{NPPES} NPI number
    \end{itemize}
\end{frame}

    \begin{block}{\normalsize Takeaway}
        \normalsize Most hospitals had incentive to improve quality to try to avoid meaningful penalties, which were not perfectly predictable
    \end{block}
\end{frame}

\begin{frame}{Executive Teams}
    \begin{table}[ht!]
        \centering
        \begin{tabular}[t]{lccc}
        \toprule
         & Ever & Always & Never\\
         & Clinical Exec & Clinical Exec &  Clinical Exec \\
        \midrule
        Number of Executives & 4.85 & 5.25 & 3.11\\
        \addlinespace[0.2em]
        Fraction Clinical Execs & 0.20 & 0.26 & 0.0\\
        \addlinespace[0.2em]
        Has Clinical CEO & 0.25 & 0.11 & 0.0\\
        \addlinespace[0.2em]
        Has CMO & 0.53 & 0.45 & 0.15\\
        \addlinespace[0.6em]
        Number of Hospitals & 391 & 56 & 445\\
        \bottomrule
        \end{tabular}
    \end{table}
\end{frame}

    \includegraphics[width=.7\textwidth]{Objects/change_analysis_plot.pdf}

\begin{frame}{Raw Outcomes, controlling for baseline case mix index}
\centering
    \includegraphics[width=.9\textwidth]{Objects/outcomes_graph_presentation.pdf}
\end{frame}

@@

\begin{transitionframe}
\centering \LARGE
    \textcolor{black}{Effect of Clinical Leadership on Response to P4P}
\end{transitionframe}

\begin{frame}{Estimation}\label{est_equation}
    \textbf{Compare always clinical to never clinical hospital executive teams:}

    \vspace{3mm}\large
    \begin{equation*}
    y_{ht} = \beta \text{ (clinical training x post 2012)}_{ht} + \gamma_{h} + \delta_t + \epsilon_{ht}
    \end{equation*}

    \vspace{5mm}
    \normalsize
    \begin{wideitemize}
        \item $y_{ht}$: weighted average readmission/ mortality rates
        \item Leverage the \textit{timing} of the programs
    \end{wideitemize}

    Different methods:
    \begin{itemize}
        \item No weights
        \item \textbf{``Synthetic DiD" type weights} \hyperlink{synthdid_equation}{\beamergotobutton{weighted}}
        \item Coarsened exact matching \hyperlink{match_equation}{\beamergotobutton{CEM}}
    \end{itemize}
\end{frame}

\begin{frame}{Identification}
\begin{wideitemize}
    \item Trends would have continued in a parallel manner
    \item No other 2012 event correlated with the composition of leadership team and the outcomes
    \item No other hospital characteristics that are perfectly correlated with clinical leadership are driving the results
    \item No anticipation of the incentive changes
\end{wideitemize}
\end{frame}

\begin{frame}{Readmission Rates}\label{main:readrates}
    \begin{columns}
    \column{0.5\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{Objects/read_md_nomd_synth_graph.pdf}
    \column{0.5\textwidth}
        \begin{wideitemize}
            \item Non-clinical teams reduce readmissions by .3 ppts more than clinical teams 
            \item Small relative to average readmission rate of 21
            \item Large difference relative to the average effect of these programs (1ppt)
        \end{wideitemize}

@@

        \hyperlink{int_margin}{\beamergotobutton{intensive margin}}

    \end{columns}
\end{frame}

\begin{frame}{Mortality Rates}
    \begin{columns}
    \column{0.5\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{Objects/mort_md_nomd_synth_graph.pdf}
    \column{0.5\textwidth}
        \begin{wideitemize}
            \item No statistical difference in mortality between clinical and non-clinical teams
        \end{wideitemize}
    \end{columns}
\end{frame}

@@

    \textbf{Compare both nonprofit leadership team types to the response of for-profits:}

    \vspace{3mm}\large
    \begin{equation*}
    y_{ht} = \beta \text{ (for-profit x post 2012)}_{ht} + \gamma_{h} + \delta_t + \epsilon_{ht}
    \end{equation*}

    Comparison group:
    \begin{itemize}
        \item nonprofit clinical; nonprofit non-clinical
    \end{itemize}

    Synthetic diff-in-diff type weights

@@

\begin{frame}{Readmission Rate}
    \begin{columns}
    \column{0.5\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{Objects/fp_read_md_synth_graph.pdf}
    \column{0.5\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{Objects/fp_read_nomd_synth_graph.pdf}
    \end{columns}
\end{frame}

\begin{frame}{Mortality Rate}
    \begin{columns}
    \column{0.5\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{Objects/fp_mort_md_synth_graph.pdf}
    \column{0.5\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{Objects/fp_mort_nomd_synth_graph.pdf}
    \end{columns}
\end{frame}

@@

\begin{frame}{Hypotheses}
    \textbf{Clinical leaders represent underlying $\alpha$}

    \begin{wideitemize}
        \item Timing of the clinical leader doesn't matter \item Clinical leader represents underlying objectives of the hospital
    \end{wideitemize}

    \textbf{Clinical leaders effectively change $\alpha$}

    \begin{wideitemize}
        \item Timing of the clinical leader does matter
        \item Clinical leaders manage the hospital differently
    \end{wideitemize}
\end{frame}

\begin{frame}{Estimation}
    \begin{figure}[ht!]
\begin{center}

 \begin{tabular}{| m{17em} |}
 \hline
 Signaling:\\ [0.5ex]
 \hline\hline 
 \vspace{2mm}
 Treat Group:  \hspace{15mm} Never MD \\
 \vspace{2mm}
 Comparison Group: \hspace{3mm} Ever MD  \\
 [1ex]
 \hline
 \end{tabular}
\hfil   %<---
 \begin{tabular}{|m{19em}|}
 \hline
 Managing:\\ [0.5ex]
 \hline\hline
 \vspace{2mm}
 Treat Group: \hspace{11mm} MD in 2012 \\
 \vspace{2mm}
 Comparison Group:  Ever MD (not in 2012)  \\
 [1ex]
 \hline
 \end{tabular}

\end{center}
 \end{figure}

@@

\begin{frame}{Results}
    \begin{table}[ht!]
\centering
\begin{tabular}[t]{lcccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{\underline{Weighted Avg. Readmission Rate}} & \multicolumn{2}{c}{\underline{Weighted Avg. Mortality Rate}} \\
 & (1) & (2) & (3) & (4)\\
\midrule
Signaling Effect & -0.01 &  & 0.03 & \\
 & (0.07) &  & (0.06) & \\
Managing Effect &  & 0.17 &  & 0.12\\
 &  & (0.1) &  & (0.09)\\
 &  &  &  & \\
Observations & 3,005 & 1,595 & 3,005 & 1,595\\
\bottomrule
\end{tabular}
\end{table}
\end{frame}

\begin{frame}{Conclusion}
    \textbf{We know very little about whether leaders matter in health care}

    \begin{wideitemize}
        \item In this paper, I show that non-clinically trained hospital executives internalize financial incentives on quality more than clinically trained executives
        \item Nonprofit objectives are not all created equal, and financial incentives might not be the only way to improve value in health care
    \end{wideitemize}

    \textbf{Future Work}

    \begin{itemize}
        \item Effects of mandated gender and racial diversity on hospital behavior
        \item Networks of executives over time 
        \item Are leaders correlated with hospital fraudulent behavior?
    \end{itemize}
\end{frame}

\begin{transitionframe}
\centering \Large
    Thank you!

    \textcolor{black}{hkagele@emory.edu}
\end{transitionframe}
